---
title: "full_CNV_pipeline"
author: "Lorke"
date: "12/2/2018"
output: html_document
---

## insert bash scripts ##


## Defining the copy number iteration functions

```{r copynumfun, echo=FALSE} 
# echo=FALSE - hides code, displays only output in the knitted doc

copynum.iter.iter <- function( y.mat, C, prior, max.iter=10, as.integer=TRUE, verbose=TRUE ){
  # y.mat must have one row for each stringency and one column for each gene
  # prior - initial estimate of copy numbers for each ref. gene
  gamma.hat <- prior
  gamma.old <- rep( 1000, length( prior ) )
  if( verbose ) cat( "copynum.iter.iter:\n" )
  # tol(erance) minimum difference between consecutive copy number estimates after each iteration step
  # if the diff. between 2 consecutive cn estimates is smaller than this, we stop iterating
  tol <- 0.01
  
  niter <- 0
  while( (max( abs( gamma.hat - gamma.old ) ) > tol) & (niter<max.iter) ){
    gamma.old <- gamma.hat
    lst <- copynum.iter( y.mat, C, prior, max.iter, as.integer, verbose )
    gamma.hat <- lst$Gamma
    phi.hat <- lst$Phi
    prior <- round( gamma.hat )
    niter <- niter + 1
    if( verbose ) cat( " iteration.iteration", niter, "\n" )
  }
  return( list( Gamma=gamma.hat, Phi=phi.hat ) )
}


copynum.iter <- function( y.mat, C, prior, max.iter=10, as.integer=TRUE, verbose=TRUE ){
# y.mat must have one row for each stringency and one column for each gene
  dd <- dim( y.mat )
  phi.hat <- rep( 1, dd[1] )
  gamma.hat <- prior
  gamma.old <- rep( 1000, dd[2] )
  if( verbose ) cat( "copynum.iter:\n" )
  tol <- 0.01
  
  niter <- 0
  while( (max( abs( gamma.hat - gamma.old ) ) > tol) & (niter<max.iter) ){
    gamma.old <- gamma.hat
    g.mat <- matrix( rep( gamma.hat, dd[1] ), nrow=dd[1], byrow=T )
    idx <- which( g.mat <= 0.5, arr.ind=TRUE )
    g.mat[idx] <- NA
    # phi.hat - normalized average number of observed ref. gene hits for each cutoff for this species
    # aka estimated stringency factor Fs with a hat (^)
    # assumed to be equal for all ref. genes per cutoff
    phi.hat <- rowMeans( y.mat/g.mat, na.rm=TRUE )/C
    # phi.hat replicated n times (n of ref. genes for this sp.)
    p.mat <- matrix( rep( phi.hat, dd[2] ), nrow=dd[1], byrow=F ) 
    # gamma.hat - estimated copy number for all kept ref. genes for this sp.
    # across all cutoffs, Cg with a hat (^)
    gamma.hat <- colMeans( y.mat/p.mat )/C   
    if( as.integer ) gamma.hat <- round( gamma.hat )
    niter <- niter + 1
    if( verbose ) cat( " iteration", niter, "\n" )
  }
  return( list( Gamma=gamma.hat, Phi=phi.hat ) )
}


copynum <- function( y, ref.mat, ref.gamma, as.integer=TRUE ){
  dd <- dim( ref.mat )
  idx <- which( ref.mat == 0, arr.ind=T )
  ref.mat[idx] <- 1
  y.mat <- matrix( rep( y, dd[2] ), nrow=dd[1], byrow=F )
  g.mat <- matrix( rep( ref.gamma, dd[1] ), nrow=dd[1], byrow=T )
  gamma.hat <- g.mat * y.mat/ref.mat
  if( as.integer ) gamma.hat <- round( gamma.hat )
  gamma.hat[which(gamma.hat==Inf,arr.ind=T)] <- NA
  return( gamma.hat )
}


```

## Generating table with tblastn hits for each species 

```{r hits_table_generator, echo=TRUE}

# Create table with hits for each ref. marker for each e-value cutoff (-4, -6, -8, -10)

b.bear <- list(name = "brown_bear", G.size = 2110508336, N.reads = 1121237882)
panda <- list(name = "giant_panda", G.size = 2405352861, N.reads = 877225626)
p.bear <- list(name = "polar_bear", G.size = 2192934624, N.reads = 891832240)
sun_bear <- list(name = "sun_bear", G.size = 2192934796, N.reads = 691832240)

all_species <- list(b.bear, panda, p.bear, sun_bear)


get_species_files <- function(species_name) {
  species_path = paste0("/home/derezanin/temp_storage/bears/", species_name, "/tblastn_hits/e2_hits_80aa")
  files <- list.files(path = species_path, full.names = TRUE)
  return(files)
}

get_gene_e_name <- function(gene_name, e) {
  return(paste0(gene_name, "_e_", e))
}

get_gene_names <- function() {
  first_species = all_species[[1]]
  files = get_species_files(first_species$name)
  gene_names <- c()
  for ( f in files ) {
    data <- read.csv(f, header=FALSE, sep="\t")
    gene_name <- strsplit(data[1,1], "_")[[1]][1]
    gene_names <- c(gene_names, gene_name)
  }
  
  return(gene_names)
}

initialize_hits_table <- function(all_species) {
  
  num_columns = length(all_species)
  num_genes = 19
  num_e_cutoffs = 4
  num_rows = num_genes * num_e_cutoffs

  gene_e_names = c()
  for (gene_name in get_gene_names()) {
    e_cutoffs <- c(4,6,8,10)
    for ( e in e_cutoffs ) {
      gene_e_name <- get_gene_e_name(gene_name, e)
      gene_e_names <- c(gene_e_names, gene_e_name)
    }
  }
  
  hits_table = matrix(0, nrow = num_rows, ncol = num_columns)
  colnames(hits_table) <- sapply(all_species, function(x) x[["name"]])
  rownames(hits_table) <- gene_e_names
  
  return(hits_table)
}

fill_hits <- function(all_species, hits) {
  for (s in all_species) {
    files <- get_species_files(s$name)
    
    for ( f in files ) {
      data <- read.csv(f, header=FALSE, sep="\t")
      data.e <- data[,11]
      gene_name <- strsplit(data[1,1], "_")[[1]][1]
      
      e_cutoffs <- c(4,6,8,10)
      for ( e in e_cutoffs ) {
        num_hits <- length(data.e[data.e<=10^-e])
        gene_e_name <- get_gene_e_name(gene_name, e)
        hits[gene_e_name, s$name] <- num_hits
      }
    } 
  }
  
  return(hits)
}

empty_hits <- initialize_hits_table(all_species)
gene_names <- get_gene_names()
hits <- fill_hits(all_species, empty_hits)

```



## Script 1 of the procedure.

Here we first decide which reference markers are possible to use for estimating copy numbers.

* Step 1 - Sanity check on read coverage.
* Step 2 - Check on effect of BLAST-cutoff.
* Step 3 - Final check if copy number estimates are not 0.


```{r script1_ref_markers_mammals}

options( stringsAsFactors = FALSE )
#source( "copynumfun.R" )


### Loads data and print a summary
cat( "Loading data...\n" )
#source( "hits_table_generator.R" )
uref <- gene_names
ref.hits <- hits

# marker length in bp
marker.length <- 80*3                            
# G.size <- colMeans( genome.size )        
G.size <- sapply(all_species, function(x) x[["G.size"]])   # assembly size estimated from k=31-91, s=2 (kmergenie)
n.species <- ncol( ref.hits )                       # number of species observed 
n.ref <- length( uref )                          # number of reference markers/genes
N.reads <- sapply(all_species, function(x) x[["N.reads"]])   # N of concatenated raw PE reads (~100bp)
cuts <- c(-4,-6,-8,-10)
cat( "   we have data for", n.ref, "reference markers\n" )
cat( "   we have data for", n.species, "species\n" )
cat( "   reference markers are", marker.length, "bases long\n" )
x <- sapply( 1:n.species, function(i){
  cat(colnames(ref.hits)[i], "has", G.size[i],"basepairs and",N.reads[i],"reads\n")
} )


### First data sanity check - checking for too low coverage
### Some reference markers may, for some reason, have extremely low coverage
### We would like to exclude these since they may affect all other estimates
### of copy numbers later
low.cov <- matrix( 0, nrow=n.species, ncol=n.ref )
rownames( low.cov ) <- colnames( ref.hits )
colnames( low.cov ) <- uref
coverage <- (N.reads*marker.length)/G.size    # This is the expected coverage per species
for( i in 1:n.ref ){
  cat( "Reference ", uref[i], "\n" )
  
  idd <- seq( 1, nrow(ref.hits), 4 )                          # hits under cutoff -4
  ohits_cutoff5 <- ref.hits[idd[i],]                          # observed number of hits for reference gene uref[i] at cutoff -4
  normalized_hits <- ohits_cutoff5/coverage                   # number of hits per unit of coverage (~percentage of hits compared to coverage)
  phi1 <- mean( normalized_hits )                             # very rough estimate of cutoff-factor Fs
  ehits1 <- coverage*phi1                                     # expected number of hits based on phi1
  # chi- square - for diff. between observed and estimated hits is high
  rr1 <- sign( ohits_cutoff5 - (ehits1-3*sqrt(ehits1)) )      # if rr1 is negative the coverage is very low
 
  
  idd <- seq( 2, nrow(ref.hits), 4 )                          # hits under cutoff -6
  phi2 <- mean( ref.hits[idd[i],]/coverage )                  # very rough estimate of cutoff-factor
  ehits2 <- coverage*phi2                                     # expected number of hits based on phi2
  rr2 <- sign( ref.hits[idd[i],] - (ehits2-3*sqrt(ehits2)) )  # if rr2 is negative the coverage is very low

  idd <- seq( 3, nrow(ref.hits), 4 )                          # hits under cutoff -8
  phi3 <- mean( ref.hits[idd[i],]/coverage )                  
  ehits3 <- coverage*phi3                                     # expected number of hits based on phi3 
  rr3 <- sign( ref.hits[idd[i],] - (ehits3-3*sqrt(ehits3)) )

  idd <- seq( 4, nrow(ref.hits), 4 )                          # hits under cutoff -10
  phi4 <- mean( ref.hits[idd[i],]/coverage )
  ehits4 <- coverage*phi4
  rr4 <- sign( ref.hits[idd[i],] - (ehits4-3*sqrt(ehits4)) )
  
  rmat <- matrix( c(rr1,rr2,rr3,rr4), ncol=4, byrow=F )       # The rr1,...,rr4 values
  low.cov[,i] <- rowSums( rmat )                              # summing the signs, i.e. need 3 out of 4 negative to
}                                                             # get a negative sum
ref.keep <- (low.cov >= 0 )
r.disc <- rowSums(!ref.keep)
f.disc <- colSums( !ref.keep )                                # number of species for which each ref.gene was discadred   
cat( "Number of discarded reference markers per species:\n" )
x <- sapply( 1:n.species, function(i){
  cat(rownames(ref.keep)[i], "discards",r.disc[i],"reference markers\n")
} )
cat( "Number of discarded species per reference marker:\n" )
x <- sapply( 1:n.ref, function(i){
  cat(colnames(ref.keep)[i], "is discarded in",f.disc[i],"species\n")
} )


### Next, checking which reference genes have read-hits that
### cannot be explained by the cutoff-model.
### In short, we expect that as the blast-cutoff is made stricter, the
### number of hits should decrease in a smooth way.
### Here we look for cases where this is clearly violated
mse <- matrix( 0, nrow=n.species, ncol=n.ref )       # mse represents residual variance (S2g) for a marker 
colnames( mse ) <- uref
rownames( mse ) <- colnames( ref.hits )
for( ss in 1:n.species ){
  cat( rownames(mse)[ss], ":\n" )
  # same as C <- coverage[ss] 
  C <- (N.reads[ss]*marker.length)/G.size[ss]     # coverage for this species
  idx.in <- which( ref.keep[ss,] )                # indices of ref. genes used for this species
  n.in <- length( idx.in )                        # number of ref. genes used for this species
  # all observed hits for this species, columns = ref.genes, rows = 4 cutoffs
  y.mat <- matrix( ref.hits[,ss], nrow=4, ncol=n.ref, byrow=F ) 
  y.mat <- y.mat[,idx.in]                         # hits for kept ref. genes for this species
  lst <- copynum.iter.iter( y.mat, C, rep( 1, n.in ), as.integer=FALSE )
  
  g.mat <- matrix( rep( lst$Gamma, 4 ), nrow=4, byrow=TRUE )
  p.mat <- matrix( rep( lst$Phi, n.in ), ncol=n.in, byrow=FALSE )
  # estimated number of ref. gene hits for this species for all 4 cutoffs
  y.hat <- C * g.mat * p.mat
  # difference between observed and estimated ref.gene hits 
  r.mat <- y.mat - y.hat
  # calculation of residual variance S^2g (formula 7. in SI) for each gene
  mse[ss,idx.in] <- apply( r.mat, 2, function(x){sum(x^2)/(4-1)} )
  # range returns min and max of given arguments
  rr <- range( c( as.vector( y.mat ), as.vector(y.hat) ) )
  plot( rr, rr, type="l", col="red", main=rownames(mse)[ss], 
        xlab="Observed number of reads", ylab="Predicted number of reads" )
  points( y.mat, y.hat, pch=16 )
  Sys.sleep( 2 )
}
# graph for observed and estimated ref. gene hits - closer the points are to the red line, smaller the diff.


sigma2 <- mean( mse, trim=0.01 )          # sigma2 is the variance of the error term, trimmed off 1% of extreme values
lambda <- (4-1)*mse/sigma2                # if lambda is large, it indicates that residual variance 
limit <- qchisq( 0.99, df=(4-1) )         # is much larger than we expected - data fits the model poorly
ref.keep <- (lambda <= limit)&ref.keep    # marker kept if lambda is lower than 99% quantile of the chi-square 
                                          # distribution with 3 degrees of freedom

cat( "Number of discarded reference markers per species:\n" )
x <- sapply( 1:n.species, function(i){
  cat(rownames(ref.keep)[i], "discards",rowSums(!ref.keep)[i],"reference markers\n")
} )
cat( "Number of discarded species per reference marker:\n" )
x <- sapply( 1:n.ref, function(i){
  cat(colnames(ref.keep)[i], "is discarded in",colSums( !ref.keep )[i],"species\n")
} )


### Third, eliminating reference genes that still estimate 
### to 0 copies, as these will blow the other copy number 
### estimates sky high later...
ref.hat <- matrix( 0, nrow= n.species, ncol=n.ref )
for( ss in 1:n.species){
  cat( rownames(ref.keep)[ss], ":\n" )
  C <- (N.reads[ss]*marker.length)/G.size[ss]
  idx.in <- which( ref.keep[ss,] )
  n.in <- length( idx.in )
  y.mat <- matrix( ref.hits[,ss], nrow=4, ncol=n.ref, byrow=F )
  y.mat <- y.mat[,idx.in]
  lst <- copynum.iter.iter( y.mat, C, rep( 1, n.in ), as.integer=FALSE )
  ref.hat[ss,idx.in] <- lst$Gamma
}
ref.keep <- (round(ref.hat) > 0)&ref.keep
cat( "Number of discarded reference markers per species:\n" )
x <- sapply( 1:n.species, function(i){
  cat(rownames(ref.keep)[i], "discards",rowSums(!ref.keep)[i],"reference markers\n")
} )
cat( "Number of species per discarded reference marker:\n" )
x <- sapply( 1:n.ref, function(i){
  cat(colnames(ref.keep)[i], "is discarded in",colSums( !ref.keep )[i],"species\n")
} )

save( ref.keep, file="res/ref_keep.RData" )

```


## Script 2 of the procedure.

You need to run script1 prior to this.
Here we estimate the reference markers copy numbers.

* Step 1 - Point estimates
* Step 2 - Bootstrapping 


```{r script2_ref_bootstraps_mammals}

options( stringsAsFactors = FALSE )
#source( "copynumfun.R" )
#source( "hits_table_generator.R" )

### Loads data and print summary
cat( "Loading data...\n" )
load( "res/ref_keep.RData" )
uref <- gene_names
ref.hits <- hits
n.species <- ncol( ref.hits )                         # N of species observed 
n.ref <- length( uref )                            # N of reference markers/genes
cuts <- c(-4,-6,-8,-10)
cat( "   we have data for", n.ref, "reference markers\n" )
cat( "   we have data for", n.species, "species\n" )
cat( "   reference markers are", marker.length, "bases long\n" )
x <- sapply( 1:n.species, function(i){
  cat(colnames(ref.hits)[i], "has", G.size[i],"basepairs and",N.reads[i],"reads\n")
} )


### Point estimates of reference marker copy number in each species
### are based on read-counts and ALL reference markers for the specific species.
### The matrix ref.point contains one row for each species(__ rows)
### and one column for each marker (19), but since some
### markers have been excluded for the various species (see script1)
### there are some NA in the matrix.

# point estimation uses sample data to calculate a single value to serve as a best guess of copy numbers for sp.
ref.point <- matrix( NA, nrow=ncol(ref.hits), ncol=n.ref )
rownames( ref.point ) <- names( G.size )
colnames( ref.point ) <- uref
for( ss in 1:n.species ){
  cat( rownames(ref.point)[ss], ":\n" )
  # same as C <- coverage[ss] 
  C <- (N.reads[ss]*marker.length)/G.size[ss]         # coverage for this species
  idx.in <- which( ref.keep[ss,] )                    # indices of ref. genes used for this species
  n.in <- length( idx.in )                            # number of ref. genes used for this sp.
  # all observed hits for this sp., columns = ref.genes, rows = 4 cutoffs
  y.mat <- matrix( ref.hits[,ss], nrow=4, ncol=n.ref, byrow=F )
  y.mat <- y.mat[,idx.in]                             # hits for kept ref. genes for this sp.
  lst <- copynum.iter.iter( y.mat, C, rep( 1, n.in ), as.integer=FALSE )
  # estimated copy number for all kept ref. genes for this sp.
  ref.point[ss,idx.in] <- lst$Gamma
}
save( ref.point, file="res/ref_point.RData" )
boxplot(ref.point,las=2,pch=16,ylab="Reference marker copy number")



### The bootstrap procedure

# The difference to the point estimate above is that
# estimates are based on a bootstrap-sample of the reference
# markers for current species (not all markers), and this is repeated 1000 times.
### When estimating the copy number for marker g, this marker must of course
# be included, but the OTHER markers are bootstrapped 1000 times, and this
# is repeated for every marker for each species.
### The 3-dimensional array ref.boot stores the results.

# phi.hat - normalized average number of observed ref. gene hits for each cutoff for this sp.
# aka estimated stringency factor Fs with a hat (^)
phi.hat <- matrix( -1, nrow=4, ncol=n.species )
rownames( phi.hat ) <- paste( "log10(E)=", cuts, sep="" ) # gives names to stringency rows 
colnames( phi.hat ) <- colnames( ref.hits )
N.boot <- 1000
# 3-dim array
ref.boot <- array( NA, dim=c(n.species, n.ref, N.boot), dimnames=list( Species=names(G.size), REF=uref, Boot=1:N.boot ) )
for( ss in 1:n.species ){
  cat( rownames(ref.point)[ss] )
  C <- (N.reads[ss]*marker.length)/G.size[ss]
  y.mat <- matrix( ref.hits[,ss], nrow=4, ncol=n.ref, byrow=F )
  idx.in <- which( ref.keep[ss,] )
  y.mat <- y.mat[,idx.in]
  n.in <- length( idx.in )
  for( g in 1:n.in ){
    idb <- which( 1:n.in != g )    # randomly sampled g-1 genes from the ref.keep for this sp.
    for( b in 1:N.boot ){
      # indices of sampled g-1 genes with replacement
      idx.boot <- c( g, sample( idb, size=(n.in-1), replace=TRUE ) ) 
      y.boot <- y.mat[,idx.boot]    # randomized observed ref. hits regarding indices in idx.boot
      lst <- copynum.iter.iter( y.boot, C, rep( 1, n.in ), max.iter=10, as.integer=FALSE, verbose=FALSE )
      ref.boot[ss,idx.in[g],b] <- lst$Gamma[1]  # boot. estimate for this marker g for this sp.
    }
    cat( "." )
  }
  cat( "\n" )
}

save( ref.boot, file="res/ref_boot.RData" )
boxplot(t(ref.boot[3,,]), ylim=c(0,4), las=2,ylab="Reference marker copy number")
abline(h=1)

```




